Supplementary material to the paper 'A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-05-10 analysis/paper/paper.Rmd
Minor edits and adding some references

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating that is separate from both radiometric and typological methods and which is widely applied. However, while the potentially immense benefits of an additional source of temporal data is undeniable, this also comes with a series of uncertainties which have been under limited scrutiny compared to those related to more ubiquitous dating techniques in archaeology. This is due to the geographical contingency and thus relative rarity of the method. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963].

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: true
institute:
  - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data in addition to radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more ubiqutous dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. This is done by means of Monte Carlo simulation, which is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous evaluations of this relationship, indicting that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

The post-glacial relative sea-level fall that characterises large areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout prehistory, but also from the fact that if archaeological phenomena were situated close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method, also called shoreline dating, has long history of use in the region and is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, grave cairns, various harbour and sea-side constructions and, as is the focus of this study, Stone Age sites [e.g. @akerlund1996; @bjerck2005; @loken1977; @nordqvist1995; @sognnes2003; @tallavaara2020; @schmitt2009; @wikell2009].

@@

The method of shoreline dating has been met with scepticism as related to the fundamental premise that most sites would have been consistently shore-bound, been characterised as a relative dating method for sites located at different elevations within a constrained geographical area, or been argued to offer no more than an earliest possible date for when a site could have been in use [see review by @nordqvist1999]. The most common application in Norway has arguably been to use shoreline dating to provide an approximate date for the occupation of the sites, often in combination with other dating methods [see for example chapters in @glorstad2002; @glorstad2003; @glorstad2004; @jaksland2001; @jaksland2012; @jaksland2012a; @jaksland2014a; @reitan2014l; @melvold2014b; @reitan2018c; @solheim2017b and below]. Recently the method has also been used independently to date a larger number sites to get a general impression of site frequency over time. This is done by aggregating point estimates of shoreline dates in 100, 200 or 500 year bins [@breivik2014; @breivik2018a; @solheim2018; @fossum2020; @mjærum2022; @nielsen2021b; see also @jorgensen2020; @tallavaara2020]. In his review, @nordqvist1999 argues that there can be little doubt concerning the general applicability of the method -- what is less clear is the level of reliability and chronological resolution that it can offer [see also @johansen1963; @johansen1997].

### 2022-05-10 analysis/supplementary-materials/supplementary.Rmd
Minor edits and adding some references

Stone Age sites for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan [@sjursjeike], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@østmo1975]. These sites are therefore not included in the analysis. Furthermore, @nielsen2021b has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic. However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.

@@

Rognlien was originally excavated in [@ingstad1970]. The site limit has been manually defined based on the description given in the report and the spatial geometry of the site as defined in the database Askeladden. The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970].The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site. Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm. Steinar Solheim, also provided in @stokke2017, 24]. Given the 0m distance that the older dates give here, however, there does not appear to be grounds on which to reject the older dates based on sea-level change.

@@

<!-- @nyland2012b -->
<!-- @danielsen2018 -->
<!-- @koxvold2018 -->
<!-- @solheim2017 -->
<!-- @granum2018 -->
<!-- @reitan2014l -->
<!-- @reitan2014f -->
<!-- @carrasco2014 -->
<!-- @carrasco2014 -->
<!-- @fossum2014b -->
<!-- @fossum2014c -->
<!-- @koxvold2017l -->
<!-- @koxvold2017n -->
<!-- @havstein2017a -->
<!-- @havstein2017 -->
<!-- @koxvold2017h -->
<!-- @eigeland2017a -->
<!-- @koxvold2013 -->
<!-- @koxvold2017d -->
<!-- @koxvold2017e -->
<!-- @koxvold2016 -->
<!-- @koxvold2017m -->
<!-- @koxvold2017f -->
<!-- @koxvold2017g -->
<!-- @koxvold2020a -->
<!-- @stokke2018a -->
<!-- @reitan2018a -->
<!-- @stokke2018a -->
<!-- @nielsen2017 -->
<!-- @nielsen2017 -->
<!-- @darmark2018c -->
<!-- @viken2018a -->
<!-- @darmark2018b -->
<!-- @eymundsson2014 -->
<!-- @eggen2014b -->
<!-- @reitan2014h -->
<!-- @mansrud2018 -->
<!-- @persson2008 -->
<!-- @eigeland2014a -->
<!-- @eigeland2014 -->
<!-- @fossum2014a -->
<!-- @olsen2013a -->
<!-- @koxvold2013a -->
<!-- @mansrud2013c -->
<!-- @hårstad2021 -->
<!-- @berge2012 -->
<!-- @schallerahrberg2012 -->
<!-- @nyland2012 -->
<!-- @amundsen2012 -->
<!-- @nyland2012a -->
<!-- @amundsen2012a -->
<!-- @jaksland2012 -->
<!-- @jaksland2012a -->
@carrasco2015
@carrasco2015
@persson2014a
@hårstad2021b
@koxvold2020
@viken2018
@darmark2018a
@darmark2018
@darmark2018
@nielsen2017a
@ekstrand2013

>>>>

Stone Age sites for which no spatial data was readily available are Frebergsvik A--C [@mikkelsen1975], Nedre Holtan [@sjurseike1991; @glorstad1998], Rugtvedt [@odgaard1994], Tangen [@glorstad2005], and Torsrød [@ostmo1976]. These sites are therefore not included in the analysis. Furthermore, @nielsen2021b has recently interpreted features from the otherwise younger sites Bratsberg and Larønningen as possibly related to farming activity in the Early Neolithic. However, given the limited number of features, the lack of any artefacts directly related to these, and the therefore somewhat speculative nature of this suggestion, the sites are kept out of the analysis here.

@@

The first archaeological finds from Rognlien were done in the 1880s and, following some smaller investigations, was excavated by @ingstad1970 in 1957. The site limit has been manually defined based on the description given in the publication and the spatial geometry of the site as defined in the database Askeladden. The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970, 26]. The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site. Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm. Steinar Solheim, also provided in @stokke2017, 24]. Given the 0m distance that the older dates give here, however, there does not appear to be grounds on which to reject the older dates based on sea-level change.

@@

<!-- @nyland2012b -->
<!-- @danielsen2018 -->
<!-- @koxvold2018 -->
<!-- @solheim2017 -->
<!-- @granum2018 -->
<!-- @reitan2014l -->
<!-- @reitan2014f -->
<!-- @carrasco2014 -->
<!-- @carrasco2014 -->
<!-- @fossum2014b -->
<!-- @fossum2014c -->
<!-- @koxvold2017l -->
<!-- @koxvold2017n -->
<!-- @havstein2017a -->
<!-- @havstein2017 -->
<!-- @koxvold2017h -->
<!-- @eigeland2017a -->
<!-- @koxvold2013 -->
<!-- @koxvold2017d -->
<!-- @koxvold2017e -->
<!-- @koxvold2016 -->
<!-- @koxvold2017m -->
<!-- @koxvold2017f -->
<!-- @koxvold2017g -->
<!-- @koxvold2020a -->
<!-- @stokke2018a -->
<!-- @reitan2018a -->
<!-- @stokke2018a -->
<!-- @nielsen2017 -->
<!-- @nielsen2017 -->
<!-- @darmark2018c -->
<!-- @viken2018a -->
<!-- @darmark2018b -->
<!-- @eymundsson2014 -->
<!-- @eggen2014b -->
<!-- @reitan2014h -->
<!-- @mansrud2018 -->
<!-- @persson2008 -->
<!-- @eigeland2014a -->
<!-- @eigeland2014 -->
<!-- @fossum2014a -->
<!-- @olsen2013a -->
<!-- @koxvold2013a -->
<!-- @mansrud2013c -->
<!-- @hårstad2021 -->
<!-- @berge2012 -->
<!-- @schallerahrberg2012 -->
<!-- @nyland2012 -->
<!-- @amundsen2012 -->
<!-- @nyland2012a -->
<!-- @amundsen2012a -->
<!-- @jaksland2012 -->
<!-- @jaksland2012a -->
<!-- @carrasco2015 -->
<!-- @carrasco2015 -->
<!-- @persson2014a -->
<!-- @hårstad2021b -->
<!-- @koxvold2020 -->
<!-- @viken2018 -->
<!-- @darmark2018a -->
<!-- @darmark2018 -->
<!-- @darmark2018 -->
<!-- @nielsen2017a -->
<!-- @ekstrand2013 -->
<!-- @ekstrand2013 -->
<!-- @ekstrand2013 -->
<!-- @amundsen2012b -->
<!-- @fossum2014 -->
<!-- @fossum2017c -->
<!-- @mansrud2016 -->
<!-- @koxvold2017j -->
<!-- @fossum2017d -->
<!-- @eggen2014 -->
<!-- @melvold2014a -->
<!-- @stokke2017 -->
<!-- @stokke2017 -->
<!-- @koxvold2017c -->
<!-- @koxvold2017b -->
<!-- @koxvold2017a -->
<!-- @reitan2021 -->
<!-- @reitan2021 -->
<!-- @reitan2014i -->
